
---
title: "Standardization of a Gaussian Markov Random Field (GMRF)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{standardize_GMRF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
# Load necessary libraries
library(scaleGMRF)
library(INLA)
```

## Introduction

The function `standardize_GMRF()` standardizes a **Gaussian Markov Random Field (GMRF)** and returns the elements needed to specify the standardized effect in the **INLA** framework. This is the main function in the `scaleGMRF` package.

Standardization is crucial to ensure that the variance parameters of the effects are correctly scaled to represent the variance contributions of their corresponding effects. This is fundamental for sensible prior specification of variance parameters, specifically when complex variance partitioning-based priors are used.

## Theory

Consider a generic effect for a covariate $X$:

$$f(X)=\mathbf{D}(X)\mathbf{u}$$
where the basis $\boldsymbol{D}(X)=[d_1(X),...,d_K(X)]$ is a row vector containing $K$ known functions evaluated at $X$ and $\boldsymbol{u}$ is a set of random coefficients.

Latent Gaussian Models assume for all effects that $\boldsymbol{u}$ is Normally-distributed. Specifically:

$$ \mathbf{u} \sim N(\mathbf{0},\sigma^2\mathbf{Q}^*)$$
where the structure matrix $\boldsymbol{Q}$ is considered known, and the scalar $\sigma^2$ is the variance parameter. 

Therefore, if we consider any given $\boldsymbol{x}$ being a column vector of realizations from $X\sim \pi(x)$, then $f(\boldsymbol{x})$ is a Gaussian Markov Random Field (GMRF).

In order to obtain a meaningful interpretation for the variance parameter $\sigma^2$ for which a prior specification is required, it is necessary to "standardize" all effects so that $\sigma^2_j$  $\approx
    Var[f_j(X_j)]$, i.e. each variance parameter approximates the variance contribution of their corresponding effect.
    
Specifically, we suggest to require:

$$\sigma^2=E_{\mathbf{u}|\sigma^2}\{Var_{X}[f(X)|\mathbf{u}]\}$$

for \textbf{fixed effects}, i.e. when the focus is on the finite-population inference;

$$\sigma^2=Var_{X,\mathbf{u}}[f(X)|\sigma^2]$$

for random effects, i.e. when the focus is on the superpopulation inference. See Ferrari \& Ventrucci for thorough motivation.

In order to ensure this requirement, we propose the following **standardization procedure**:


1. \textbf{0-mean constraint for fixed effects}


Imposing $E_X[f(X)]=0$ removes the uncertainty around the process mean and ensures the equality between the requirements for fixed and random effects:
\begin{gather*}
E_X[f(X)]=0\implies\\
      Var_{X,\mathbf{u}}[f(X)|\sigma^2]= E_{\mathbf{u}|\sigma^2}\{Var_{X}[f(X)|\mathbf{u}]\}
\end{gather*}
In practice, the constraint is equal to:

$$\boldsymbol{A}\mathbf{u}=0 \text{ where }\boldsymbol{A}=E_X[D(X)]$$

for $K>1$. If $K=1$, the single basis function must be changed to $d_1(X)-E_X[d_1(X)]$. Note that the constraint modifies the precision matrix $\boldsymbol{Q}$.

1. \textbf{Scaling} (all effects)

Similar to the scaling procedure by S{\o}rbye \& Rue {\color{cadmiumgreen!100}$[4]$}, we propose to scale all effects $f(X)/\sqrt{C}$ where the constant $C$ is equal to:

$$C=E_X\{Var_{\mathbf{u}|X}[f(X)|\sigma^2=1]\}$$

The scaling step ensures that $\sigma^2=Var_{X,\mathbf{u}}[f(X)|\sigma^2]\}$. See Ferrari \& Ventrucci (2025) for further details, examples, and practical implications.

We denote the standardized effect by:

$$\widetilde{f}(X)=\widetilde{\mathbf{D}}(X)\widetilde{\mathbf{u}} \text{  where}$$


$$\widetilde{\mathbf{u}}\sim N(\mathbf{0},\sigma^2\widetilde{\mathbf{Q}^*})$$

$\widetilde{\mathbf{u}}$ will be also subject to $\boldsymbol{A}\widetilde{\mathbf{u}}=0$ if the effect is fixed. $\widetilde{\mathbf{D}}$ and $\widetilde{\mathbf{Q}}$ may or may not be equal to the original matrices, according to whether or not the effect already satisfied the requirements in terms of $\sigma^2$.

## Function Overview

```{r, eval=FALSE}
standardize_GMRF(Q, D = NULL, rank_def = NULL, fixed = FALSE, scale_Q = TRUE, plot_check = FALSE, n_sim = 100)
```

The `standardize_GMRF()` function implements the standardization procedure for a generic effect $f(X)$. Before using the function, the user must have specified the following elements of $f(X)$:

- $\boldsymbol{D}(X)=[d_1(X),...,d_K(X)]$: the $K$ basis functions
- $\boldsymbol{Q}$: the $K\times K$ precision matrix on the coefficients
- $\pi(x)$: the probability distribution of $X$ 
- whether the effect is *fixed* or *random*

These elements are specified in the function with the following arguments:

- \texttt{Q}: A $K\times K$ matrix representing $\boldsymbol{Q}$. The optional argument `rank_def` can be used to specify the rank deficiency of $\boldsymbol{Q}$ and therefore speed up computation.

- `D`: A $N \times K$ matrix representing $\boldsymbol{D}(\boldsymbol{x})$
 where $\boldsymbol{x}=[x_1,...,x_N]^T$ must be a large i.i.d. sample from $\pi(x)$.
 
- `fixed`: A logical value that specifies if the effect should be treated as fixed. By default, the function treats the effect as random.

The argument `D` is optional as the default choice covers the simplest and common scenario where $d_k(X)=I(X=k)$ and $p(X=k)=1/K$.

The output of `standardize_GMRF()` is a list of 4 elements, namely:

- `Q`: a matrix representing $\widetilde{\boldsymbol{Q}}$ (or $\widetilde{\boldsymbol{Q}}/C$ when `scale_Q=FALSE`)
- `D`: a matrix representing $\widetilde{\boldsymbol{D}}(\boldsymbol{x})$. Equal to the input `D` unless $K=1$.
- `A`: a matrix with 1 row, representing $\boldsymbol{A}$. `NULL` for `fixed=FALSE`.
- `C`: a positive number, representing the scaling constant $C$

The logical argument `scale_Q` is a logical value that determines whether the scaling step should be applied. By default, it is obviously set to `TRUE`, but the option is useful to compare effects before and after the scaling to assess the practical impact of the procedure. 

## Diagnostic plot

The logical argument `plot_check` indicates if a diagnostic plot should be generated to check the standardization. The diagnostic plot is based on the simulation of a number of realizations `n_sim` (by default, 100) of $f(\boldsymbol{x})$ when $\sigma^2=1$. The plot changes for fixed and random effects.

- *Fixed effects*. Since the requirement for fixed effects is that $\sigma^2=E_{\mathbf{u}|\sigma^2}\{Var_{X}[f(X)|\mathbf{u}]\}$, we want to check that when $\sigma^2=1$, then the distribution of $Var_{X}[f(X)|\mathbf{u}]$ for multiple realizations of $\mathbf{u}$ is concentrated around 1. To do so, we plot the histogram of the *within-realization variance*, i.e.:
$Var_{i=1}^N[f(x_i)]$
for each realization of $f(\boldsymbol{x})$. The plot also reports the mean of the within-realization variance as a dashed black line, as well as its numerical value at the bottom: if the standardization is succesful, the mean should be close to 1.

```{r}
#Quadratic effect
x <- sort(runif(100))
basis <- matrix(x^2)
standardize_GMRF(Q = matrix(1), D = basis, fixed = TRUE, plot_check = TRUE, n_sim=10000)
```

- *Random effects*. Since the requirement for random effects is that $\sigma^2=Var_{X,\mathbf{u}}[f(X)|\sigma^2]\}$, which is also equal to $\sigma^2=E_{X}\{Var_{\mathbf{u}}[f(X)|\sigma^2]\}$ we want to check that when $\sigma^2=1$, then the distribution of $Var_{\mathbf{u}}[f(X)|\sigma^2]$ for multiple realizations of $X$ is concentrated around 1. To do so, we compute the *between-realization variance* for each value $x_i$ in $\boldsymbol{x}$. For B realizations, we compute:
$Var_{b=1}^B[f^{(b)}(x_i)]$
Since the function does not directly take as argument $\boldsymbol{x}$ but only $D(\boldsymbol{x})$, we simply plot the *between-realization variance* in the order in which the user provided. The plot also reports the mean of the between-realization variance as a dashed black line, as well as its numerical value at the bottom: if the standardization is succesful, the mean should be close to 1.

```{r}
#Random walk effect
x <- 1:20
Q_matrix <- as.matrix(spam::precmat.RW1(length(x)))
standardize_GMRF(Q = Q_matrix, plot_check = TRUE,n_sim = 10000)
```

## Using the Output in INLA

Once the GMRF is standardized using `standardize_GMRF()`, you can use the output directly in an **INLA** model to specify an effect. The key components of the output—`Q`, `D`, and `A`—are used to define a GMRF effect in INLA.


As an example, we consider a latent Gaussian model for a Normally-distributed response $Y$, where the linear predictor is a first-order random walk on $K=20$ locations:

```{r}
# Creation of the data
x <- 1:20
y <- cumsum(rnorm(20))+rnorm(20)
plot(x,y)
```

Then, we find the standardized version of a first-order random walk using `standardize_GMRF()`:

```{r}
# Standardization of the effect
Q_matrix <- as.matrix(spam::precmat.RW1(length(x)))
standardized_effect <- standardize_GMRF(Q = Q_matrix,plot_check = T,fixed=T,n_sim = 10000)
# Compare to the unstandardized version
standardize_GMRF(Q = Q_matrix,scale_Q = F, fixed=T,plot_check = T,n_sim = 10000)
```

Now, we can start specifying our INLA model. First, we create the data stack and use the basis of the standardized effect, stored in the `D` element of `standardized_effect`:
```{r}
library(INLA)

# Create the data stack for INLA
data_stack <- inla.stack(
  data = list(Y = y),       #response
  A = list(1, standardized_effect$D),  # Use the standardized basis
  effects = list(
    data.frame(intercept = rep(1, length(y))),
    list(u_rw1 = 1:ncol(standardized_effect$D))  # Random walk coefficients
  )
)
```

Secondly, we need to include the effect in the formula of the INLA call. To specify a generic GMRF effect with null mean, we need to set `model = "generic0"`; then, the standardized version of the precision matrix can be specified using the argument `Cmatrix`; finally, remember to include the 0-mean constraint on the coefficients using the element `A` if the effect is considered fixed.

```{r}
# Fit the INLA model
model <- inla(
  Y ~ -1 + intercept +  # No default intercept
    f(u_rw1, model = "generic0", constr = FALSE,
      Cmatrix = standardized_effect$Q,  # Standardized precision matrix
      extraconstr = list(
        A = standardized_effect$A,  # Mean constraint (if fixed effect)
        e = matrix(rep(0, 1))) # Constraint RHS
    ),
  family = "gaussian", data = inla.stack.data(data_stack),
  control.predictor = list(A = inla.stack.A(data_stack)),verbose=T
)

summary(model)
```


